Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-07-12 Manuscript/manuscript.Rmd
results edits

To establish the data for classification, the genome was binned into equally sized intervals defined by the data resolution of the corresponding contact matrix. As an example, for domain data obtained from a 10 kb resolution contact matrix, the genome was binned into 10 kb intervals. In order to accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. That is, for 10 kb resolution data, each uniquely identified TAD boundary was flanked on either side by 10 kilobases for a total width around each boundary point of 20 kb. A genomic bin was then labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0).

@@

where TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary (false negatives). Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. Additional performance metrics were [??? are?] reported in the supplementary materials. [??? Where is it?]

@@

@@

We analyzed publically available replicate Hi-C data sets from two cell lines including GM12878 and K562 (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at four resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. To account for the varying levels of data resolution, we developed a resolution-specific approach toward constructing TAD versus non TAD boundary regions on the linear genome, labeled as a response vector **Y** (Figure 1). First, we binned the linear genome according to the data resolution. We then flanked the boundaries on either side by the same number of basepairs as the previously constructed genomic bins. TAD boundary regions were defined as regions on the linear genome where a genomic bin overlapped with a flanked TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as the regions that did not overlap with a flanked TAD boundary ($Y = 0$).

@@

In order to identify the optimal combination of data resolution, predictor type, and re-sampling technique necessary for TAD boundary prediction, we created an ensemble framework to systematically compare performances of different model inputs (Figure 4). Each system in the framework was assumed to be composed of a random forest classifier constructed from a particular cell line ($L = \{GM12878, K562\}$). Within each cell line we evaluated 5 data resolutions ($R = \{5 kb, 10 kb, 25 kb, 50 kb, 100 kb\}$). We then considered 3 separate predictor types as the feature space ($P = \{OC, OP, Distance\}$). To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling ($B = \{None, RUS, ROS, SMOTE\}$). Each combination of the above cell line specific parameters was implemented on chromosome specific sections of the linear genome (omitting chromosome 9). Letting $|X|$ denote the cardinality [??? TBD] of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2520 ($2 \times 5 \times 3 \times 4 \times 21$) separate models. For each combination of inputs, we performed 10-fold cross-validation to tune for the number of features to consider at each node of the random forest classifier. Each optimally tuned model was then validated on the same separate cell line specific testing set for unbiased comparisons. Performances were evaluated using MCCs and aggregated across each chromosome by averaging. <!--We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.-->

@@

We first evaluated the least granular cell line specific random forest classifiers [??? Unclear], those were ones implemented using OC predictor types, without data re-sampling. We found that these models exhibited poor MCCs across all resolutions (Figure 5 A-B (red)). We then assessed models built on OP predictor types, while still not performing any data resampling. Performances were similar to that of OC types, with no clear improvement (Figure 5 C-D (red)). This indicated that the added complexity of using overlap percents from the functional genomic annotations did not affect predictive performance, regardless of cell line and resolution. However, when cell line specific classifiers were implemented using distance-type predictors and no re-sampling, we found that there was a significant improvement in MCC compared to the other two predictor types (Figure 5 E-F (red)).

We then further explored the impact of data re-sampling on model performance, across cell line and resolution. For cell line specific random forest classifiers built on balanced data, we found that SMOTE re-sampling provided the optimal MCC, consistently outperforming models built on ROS or RUS, across both predictor type and resolution (Figure 5). Specifically, the most improvement was seen when combining SMOTE re-sampling with distance-type predictors for both the GM12878 and K562 cell lines.

@@

Interestingly, we found that there was a degredation in model performance as data resolution decreased. Chromosome specific models built on 10 kb data resolutions presented the highest MCC values (Figure 5). Additionally, performances tended to be much more variable between chromosomes at lower resolutions. Moreover, this decreasing trend in performance was much more evident under the optimal combination of data level characteristics. These results imply that an optimally performing cell line specific random forest classifier <!--, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries,--> is one that is built on 10 kb resolution data, with distance-type predictors, and incorporates both random under- and over-sampling.

@@

After determining the optimal set of data level characteristics, we then aimed to compare the performances of annotation-specifc models [??? Never defined, what is this?; !!!defined above; should i mention again?], within each cell line and across resolution. In doing so, we inspected the proportion of predictive importance that was explained by each of the 3 groups of annotations (chromatin states, histone modifications, and transcription factor binding sites). We found that for cell line specific random forest models implemented on distance-type predictors, using SMOTE re-sampling, the majority of the proportional predictive importance was obtained by the functional genomic annotations classified as transcription factor binding sites (TFBS) (Figure 6). This was consistent across resolutions, suggesting that TFBS were the primary drivers of TAD boundary formation in both the GM12878 and K562 cell lines.

We further assessed the predictive capabilities of the specific sets of annotations by comparing the performances of annotation-specific random forest models implemented on the optimal set of characteristics. We found that cell line specific models built simply on TFBS outperformed other cell line and annotation-specific models at all resolutions (Figure 7). Additionally, when compared to models built on all 3 sets of functional genomic annotations, the TFBS-specific models performed similarly at the 10 kb and 25 kb resolutions. Interestingly though, at the 50 kb and 100 kb resolutions, the TFBS-specific models out performed models built on the combined sets of annotations for both cell lines. This indicated that, at lower resolutions, the inclusion of other types of annotations contributed to additional noise rather than signal.

@@

We then sought to determine which of the transcription factor binding sites were most influential in predicting TAD boundaries. We found that the cell line specific binding sites associated with SMC3, CTCF, RAD21, ZNF143, and YY1sc281 were given consistently high predictive importance, across all resolutions (Figure 7; Supplementary Figure 6). The SMC3, CTCF, and RAD21 binding sites are known factors associated with the cohesin protein complex, which is hypothesized to be what creates loops in the DNA  [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, ZNF143 and YY1 are known structural regulators of enhancer-promoter interactions that also contribute to looping of the linear genome [@weintraub2017yy1; @gonzalez2017znf143]. With the identification of these factors known to be enriched at TAD boundaries, we are able to confirm the validity of our model building process. Additionally, we were also able to identify some lesser known cell line specific transcription factors as predictive of TAD boundaries including the MYC-associated zing-finger protein (MAZ), jun-D proto-oncogene (JUND), and ELF1. These transcription factor binding sites tended to be given higher predictive importance at lower resolutions. [??? Where do you show it?] In fact, at lower resolutions, the predictive importance was found to be much more evenly dispersed across the top transcription factor binding sites, than for higher resolutions. This is again suggestive of the fact that the spatial relationship between functional annotations, specifically transcription factor binding sites, and TAD boundaries changes at different Hi-C data resolutions.

@@

![**Assessing optimally built cell line specific random forest classifiers for predicting TAD boundary formation.** (A) Comparing the proportion of total predictive importance carried by each group of functional genomic annotation across resolution. Predictive importance was averaged for models built on chromosome specific regions of the linear genome. Proportional importance was then calculated according to the annotation group of the feature space including chromatin state segmentations (CSS), histone modifications (HM), and transcription factor binding sites (TFBS). (B) Comparing models using specific groups of functional genomic annotation as the feature space in addition to a model with a feature space containing all 3 types combined (combined; purple). Error bars represent $\pm$ 1 standard deviation from the mean MCC across each chromosomal section of the linear genome. (C) Heatmaps illustrating the importance of the top 20 cell line specific transcription factors across resolution. Importances were sorted according to the row-wise average across resolution and rescaled to values between 0 and 100. Transcription factors with higher values of predictive importance are denoted in dark red and dark blue for the GM12878 and K562 cell line respectively.](Figures/prop_imps_a_comparing_annots_b_var_imps_c.png)

>>>>

To establish the data for classification, the genome was binned into equally sized intervals defined by the data resolution of the corresponding contact matrix. As an example, for domain data obtained from a 100 kb resolution contact matrix, the genome was binned into 100 kb intervals. In order to accommodate for the uncertainty in exact boundary location, the boundary points were flanked on both sides by the size of the resolution of the corresponding contact matrix. That is, for 100 kb resolution data, each uniquely identified TAD boundary was flanked on either side by 100 kilobases for a total width around each boundary point of 200 kb. A genomic bin was then labeled as having a TAD boundary within it ($Y = 1$) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary ($Y = 0$).

@@

where TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary (false negatives). Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. <!--Additional performance metrics including AUROC, accuracy, sensitivity, and specificity were [??? are?] reported in the supplementary materials. [??? Where is it?]-->

@@

[needs editting]

@@

We analyzed publically available replicate Hi-C data sets from two cell lines including GM12878 and K562 (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at 4 resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. To account for the varying levels of data resolution, we developed a resolution-specific approach toward constructing TAD versus non-TAD boundary regions on the linear genome (See Methods), labeled as a response vector **Y** (Figure 1). <!--First, we binned the linear genome according to the data resolution. We then flanked the boundaries on either side by the same number of basepairs as the previously constructed genomic bins.--> TAD boundary regions were defined as regions on the linear genome where a genomic bin overlapped with a flanked TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as the regions that did not overlap with a flanked TAD boundary ($Y = 0$).

@@

In order to identify the optimal combination of data resolution, predictor type, and re-sampling technique necessary for TAD boundary prediction, we created an ensemble framework to systematically compare performances of different model inputs (Figure 4). Each system in the framework was assumed to be composed of a random forest classifier constructed from a particular cell line (*L = {GM12878, K562}*). Within each cell line we evaluated 5 data resolutions (*R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*). We then considered 3 separate predictor types as the feature space (*P = {OC, OP, Distance}*). To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling (*B = {None, RUS, ROS, SMOTE}*). Each combination of the above cell line specific parameters was implemented on chromosome specific sections of the linear genome (omitting chromosome 9). Letting $|X|$ denote the cardinality [??? TBD] of each set, we evaluated *|L|* $\times$ *|R|* $\times$ *|P|* $\times$ *|B|* different ensemble systems, totaling 2520 (*2* $\times$ *5* $\times$ *3* $\times$ *4* $\times$ *21*) separate models. For each combination of inputs, we performed 10-fold cross-validation to tune for the number of features to consider at each node of the random forest classifier. Each optimally tuned model was then validated on the same separate cell line specific testing set for unbiased comparisons. Performances were evaluated using MCCs and aggregated across each chromosome by averaging. <!--We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.-->

@@

We first evaluated the least granular cell line specific random forest classifiers [??? Unclear; !!!discuss in meeting], those were ones implemented using OC predictor types, without data re-sampling. We found that these models exhibited poor MCCs across all resolutions (Figure 5 A-B (red)). We then assessed models built on OP predictor types, while still not performing any data resampling. Performances were similar to that of OC types, with no visible improvement across resolution, for either cell line (Figure 5 C-D (red)). This indicated that the added complexity of using overlap percents from the functional genomic annotations did not affect predictive performance. However, when cell line specific classifiers were implemented using distance-type predictors and no re-sampling, we found that there was a significant improvement in MCC compared to the other two predictor types (Figure 5 E-F (red)). We then further explored the impact of data re-sampling on model performance, across cell line and resolution. For cell line specific random forest classifiers built on balanced data, we found that SMOTE re-sampling provided the optimal MCC, consistently outperforming models built on ROS or RUS respectively, across both predictor type and resolution (Figure 5). Specifically, the most improvement was seen when combining SMOTE re-sampling with distance-type predictors for both the GM12878 and K562 cell lines.

@@

Interestingly, we found that performance peaked at 10 kb resolution and then degraded as data resolution decreased (Figure 5). The reason for the improved performance between 5 kb and 10 kb data resolutions can be attributed to the added sparcity in the interaction frequency matrices at finer resolutions. This sparcity corresponds to inaccurate TAD calling which leads to worse model performance. The difference in performance between 5 kb and 10 kb resolutions for K562 was found to be slightly steaper, indicating that this issue of sparcity was more severe at finer resolutions for K562. Additionally, performances tended to be much more variable between chromosomes at lower resolutions. Moreover, this decreasing trend in performance was much more evident under the optimal combination of data level characteristics (SMOTE resampling with distance type predictors). These results imply that there is an ideal resolution in TAD boundary prediction, and that that an optimally performing cell line specific random forest classifier <!--, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries,--> is one that is built on 10 kb resolution data, with distance-type predictors, and incorporates both random under- and over-sampling.

@@

After determining the optimal set of data level characteristics, we then aimed to compare the performances of annotation-specifc models [??? Never defined, what is this?; !!!defined above; should i mention again?], within each cell line and across resolution. In doing so, we inspected the proportion of predictive importance that was explained by each of the 3 groups of annotations (chromatin states, histone modifications, and transcription factor binding sites). We found that for cell line specific random forest models implemented on distance-type predictors, using SMOTE re-sampling, the majority of the proportional predictive importance was obtained by the functional genomic annotations classified as transcription factor binding sites (TFBS) (Figure 6 A). This was consistent across resolutions, suggesting that TFBS were the primary drivers of TAD boundary formation in both the GM12878 and K562 cell lines.

We further assessed the predictive capabilities of the specific sets of annotations by comparing the performances of annotation-specific random forest models implemented on the optimal set of characteristics. We found that cell line specific models built simply on TFBS outperformed other cell line and annotation-specific models at all resolutions (Figure 6 B). Additionally, when compared to models built on all 3 sets of functional genomic annotations, the TFBS-specific models performed similarly at the 10 kb and 25 kb resolutions. Interestingly though, at the 50 kb and 100 kb resolutions, the TFBS-specific models out performed models built on the combined sets of annotations for both cell lines. This indicated that, at lower resolutions, the inclusion of other types of annotations contributed to additional noise rather than signal.

@@

We then sought to determine which of the transcription factor binding sites were most influential in predicting TAD boundaries. We found that the cell line specific binding sites associated with SMC3, CTCF, RAD21, and ZNF143 were given consistently high predictive importance, across all resolutions (Figure 6 C). The SMC3, CTCF, and RAD21 binding sites are known factors associated with the cohesin protein complex, which is hypothesized to be what creates loops in the DNA  [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, ZNF143 is a known structural regulators of enhancer-promoter interactions that also contribute to looping of the linear genome [@weintraub2017yy1; @gonzalez2017znf143]. With the identification of these factors known to be enriched at TAD boundaries, we are able to confirm the validity of our model building process. Additionally, we were also able to identify some lesser known cell line specific transcription factors as predictive of TAD boundaries including the MYC-associated zing-finger protein (MAZ), jun-D proto-oncogene (JUND), and ELF1. These transcription factor binding sites tended to be given higher predictive importance at lower resolutions. [??? Where do you show it?] In fact, at lower resolutions, the predictive importance was found to be much more evenly dispersed across the top transcription factor binding sites, than for higher resolutions. This is again suggestive of the fact that the spatial relationship between functional annotations, specifically transcription factor binding sites, and TAD boundaries changes at different Hi-C data resolutions.

@@

![**Assessing optimally built cell line specific random forest classifiers for predicting TAD boundary formation.** (A) Comparing the proportion of total predictive importance carried by each group of functional genomic annotation across resolution. Predictive importance was averaged for models built on chromosome specific regions of the linear genome. Proportional importance was then calculated according to the annotation group of the feature space including chromatin state segmentations (CSS), histone modifications (HM), and transcription factor binding sites (TFBS). (B) Comparing models using specific groups of functional genomic annotation as the feature space in addition to a model with a feature space containing all 3 types combined (combined; purple). Error bars represent $\pm$ 1 standard deviation from the mean MCC across each chromosomal section of the linear genome. (C) Heatmaps illustrating the importance of the top 20 cell line specific transcription factors across resolution. Importances were sorted according to the row-wise average across resolution and rescaled to values between 0 and 100. Transcription factors with higher values of predictive importance are denoted in dark red and dark blue for the GM12878 and K562 cell line respectively.](Figures/prop_imps_a_comparing_annots_b_var_imps_c_v2.png)
